---
output:
  md_document:
    variant: markdown_github
---

# Purpose

This is the solution for the Coffee Hub analysis. The approach to the question is as follows:

- Create the loader function to read the data from the csv file to fixe the encoding issues.


```{r, echo=FALSE }

rm(list = ls()) # Clean your environment:
gc() # garbage collection - It can be useful to call gc after a large object has been removed, as this may prompt R to return memory to the operating system.
library(tidyverse)
list.files('code/', full.names = T, recursive = T) %>% .[grepl('.R', .)] %>% as.list() %>% walk(~source(.))
```

# 1. Loading the data

```{r}
coffee <- load_coffee("data/Coffee/Coffee.csv")
# the load function firstly removes the encoding issues and then once its done i filter and remove any coffee item that has missing values or an NA in the observables.

# I also create a Cost variable in Rands using the exchange rate as of June 17th 2026 which was R16.17.
# lastly, i also merger the 3 reviewers feedback into 1 long review.

```


##  KEYWORD MATCHING  —  Stellenbosch student word-cloud preferences

```{r}
# This is the top words from the Stellenbosch survey that were given in the word bubble within the key words.
# The objective is to use the how many of the reviews hit or use the same words as in the ones that were top from the survey.


student_keywords <- c(
    "sweet", "chocolate", "aroma", "mouthfeel", "finish", "structure",
    "toned", "notes", "bright", "fruit", "rich", "floral", "balanced",
    "crisp", "spice", "dark", "roasted", "cocoa", "cup", "syrupy",
    "smooth", "honey", "berry", "citrus", "caramel", "vanilla", "nutty",
    "almond", "cherry"
)


coffee <- flag_keyword_matches(coffee, student_keywords) # this creates the variable kw_hits which how many of the keywords from the survey from stellenbosch were included in the reviews. 

head(coffee)
```


# Analysis

Here i will analyse 7 key factors where i rank the coffees based of the variabke of interest.

## Top origin 

```{r}
top_origins <- function(df, n = 12, min_count = 10) {
    df %>%
        group_by(origin_1) %>%
        summarise(
            avg_rating = mean(Rating),
            avg_cost   = mean(Cost_rands),
            count      = n(),
            .groups = "drop"
        ) %>%
        filter(count >= min_count) %>%
        arrange(desc(avg_rating)) %>%
        slice_head(n = n)
}

top_origins(coffee)
```


