---
output:
  md_document:
    variant: markdown_github
---

# Purpose

This is the solution for the Coffee Hub analysis. The approach to the question is as follows:

- Create the loader function to read the data from the csv file to fixe the encoding issues.


```{r, echo=FALSE, message=FALSE, warning=FALSE }

rm(list = ls()) # Clean your environment:
gc() # garbage collection - It can be useful to call gc after a large object has been removed, as this may prompt R to return memory to the operating system.
library(tidyverse)
library(scales)
list.files('code/', full.names = T, recursive = T) %>% .[grepl('.R', .)] %>% as.list() %>% walk(~source(.))
```

# 1. Loading the data

```{r}
coffee <- load_coffee("data/Coffee/Coffee.csv")
# the load function firstly removes the encoding issues and then once its done i filter and remove any coffee item that has missing values or an NA in the observables.

# I also create a Cost variable in Rands using the exchange rate as of June 17th 2026 which was R16.17.
# lastly, i also merger the 3 reviewers feedback into 1 long review.

```


##  KEYWORD MATCHING  —  Stellenbosch student word-cloud preferences

```{r}
# This is the top words from the Stellenbosch survey that were given in the word bubble within the key words.
# The objective is to use the how many of the reviews hit or use the same words as in the ones that were top from the survey.


student_keywords <- c(
    "sweet", "chocolate", "aroma", "mouthfeel", "finish", "structure",
    "toned", "notes", "bright", "fruit", "rich", "floral", "balanced",
    "crisp", "spice", "dark", "roasted", "cocoa", "cup", "syrupy",
    "smooth", "honey", "berry", "citrus", "caramel", "vanilla", "nutty",
    "almond", "cherry"
)


coffee <- flag_keyword_matches(coffee, student_keywords) # this creates the variable kw_hits which how many of the keywords from the survey from stellenbosch were included in the reviews. 

head(coffee)
```


# Analysis

Here i will analyse 7 key factors where i rank the coffees based of the variabke of interest.

```{r}
# here i use the analyse functions for create variables that will be sueed for summary stats
origins        <- top_origins(coffee)
roast_stats    <- rating_by_roast(coffee)
value_coffees  <- best_value(coffee)
roasters       <- top_roasters(coffee)
student_picks  <- student_top_picks(coffee)
cost_stats     <- cost_summary(coffee)
locations      <- roaster_locations(coffee)
```



# visuals

```{r}
# ── Plot 1: Top Origins ──────────────────────────────────────────────────
 origins %>%
    mutate(origin_1 = fct_reorder(origin_1, avg_rating)) %>%
    ggplot(aes(x = avg_rating, y = origin_1, fill = origin_1)) +
    geom_col(show.legend = FALSE, width = 0.65) +
    geom_text(aes(label = round(avg_rating, 1)),
              hjust = -0.2, size = 3.5, colour = "#3B2314") +
    scale_fill_manual(values = coffee_palette) +
    labs(
        title = "Top Coffee Origins by Average Rating",
        x = "Average Rating", y = NULL
    ) +
    coord_cartesian(xlim = c(min(origins$avg_rating) - 0.5, NA)) +
    theme_coffee()
```


```{r}

# ── Plot 2: Rating by Roast (boxplot) ────────────────────────────────────
roast_order <- roast_stats %>% pull(roast)
 
coffee %>%
    mutate(roast = factor(roast, levels = roast_order)) %>%
    ggplot(aes(x = roast, y = Rating, fill = roast)) +
    geom_boxplot(alpha = 0.8, show.legend = FALSE, width = 0.55) +
    scale_fill_manual(values = coffee_palette) +
    labs(
        title = "Rating Distribution by Roast Type",
        x = NULL, y = "Rating"
    ) +
    theme_coffee()

```


```{r}
 
# ── Plot 3: Cost vs Rating (scatter, colour = keyword hits) ──────────────
 coffee %>%
    ggplot(aes(x = Cost_Per_100g, y = Rating, colour = kw_hits)) +
    geom_point(alpha = 0.55, size = 2) +
    scale_colour_gradient(low = "#F5DEB3", high = "#6F4E37",
                          name = "Student\nKeyword Hits") +
    labs(
        title = "Cost vs Rating (colour = student keyword match strength)",
        x = "Cost per 100g (USD)", y = "Rating"
    ) +
    theme_coffee()
```


```{r}

 roasters %>%
    mutate(roaster = fct_reorder(roaster, avg_rating)) %>%
    ggplot(aes(x = avg_rating, y = roaster, fill = roaster)) +
    geom_col(show.legend = FALSE, width = 0.6) +
    geom_text(aes(label = round(avg_rating, 1)),
              hjust = -0.2, size = 3.5, colour = "#3B2314") +
    scale_fill_manual(values = coffee_palette) +
    labs(
        title = "Top 10 Roasters (>= 5 coffees reviewed)",
        x = "Average Rating", y = NULL
    ) +
    coord_cartesian(xlim = c(min(roasters$avg_rating) - 0.5, NA)) +
    theme_coffee()
```


```{r}

med_cost <- median(coffee$Cost_Per_100g)
 
 coffee %>%
    ggplot(aes(x = Cost_Per_100g)) +
    geom_histogram(bins = 40, fill = "#6F4E37", colour = "white", alpha = 0.85) +
    geom_vline(xintercept = med_cost, linetype = "dashed",
               colour = "#D4A574", linewidth = 1) +
    annotate("text", x = med_cost + 3, y = Inf, vjust = 2,
             label = paste0("Median: $", round(med_cost, 2)),
             colour = "#3B2314", size = 4) +
    labs(
        title = "Cost Distribution of Reviewed Coffees",
        x = "Cost per 100g (USD)", y = "Number of Coffees"
    ) +
    theme_coffee()


```

