---
title: "Question 4: Netflix Streaming Analysis"
author: "Data Science 871 Practical"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 6)

# Load packages
library(tidyverse)
library(knitr)
library(kableExtra)
library(tidytext)

# Source all helper functions from the code/ folder
list.files('code/', full.names = T, recursive = T) %>%
    .[grepl('.R', .)] %>%
    as.list() %>%
    walk(~source(.))
```

# Introduction

Netflix has experienced a decline in subscribers and share price in recent times. In this report we explore the Netflix content catalogue to understand what types of content dominate the platform, how ratings and movie lengths differ across countries and genres, and what themes emerge from movie descriptions.

We work with three datasets:

- **Titles**: All Netflix shows and movies with IMDb/TMDB scores, genres, countries and run-time.
- **Movie_Info**: Netflix movies specifically, with directors, cast, age ratings and duration.
- **Credits**: Actor and character information linked to titles.

# Data Loading & Preparation

```{r load-data}
# Load all three datasets using our sourced helper functions
Titles <- load_titles("data/netflix/titles.rds")
Movie_Info <- load_movie_info("data/netflix/netflix_movies.csv")
Credits <- load_credits("data/netflix/credits.rds")
```

# Content Overview

## Movies vs Shows

```{r type-split}
# Count content types
type_counts <- Titles %>%
    count_top_n(type, n = 5)

# Simple bar chart
ggplot(type_counts, aes(x = reorder(type, n), y = n, fill = type)) +
    geom_col(show.legend = FALSE) +
    geom_text(aes(label = n), hjust = -0.2, size = 4) +
    coord_flip() +
    labs(title = "Netflix Content: Movies vs Shows",
         x = NULL, y = "Number of Titles") +
    theme_minimal() +
    scale_fill_manual(values = c("MOVIE" = "#E50914", "SHOW" = "#564d4d"))
```

- Netflix's catalogue is heavily **movie-dominated**, which makes sense given movies are cheaper to license per title than multi-season shows.

## Release Year Trends

```{r release-trend}
# Titles per year, split by type
Titles %>%
    filter(release_year >= 1990) %>%
    count(release_year, type) %>%
    ggplot(aes(x = release_year, y = n, colour = type)) +
    geom_line(linewidth = 1) +
    geom_point(size = 1.2, alpha = 0.6) +
    labs(title = "Number of Titles by Release Year (post-1990)",
         x = "Release Year", y = "Count", colour = "Type") +
    theme_minimal() +
    scale_colour_manual(values = c("MOVIE" = "#E50914", "SHOW" = "#221f1f"))
```

- Content additions ramp up sharply from around 2015, peaking in the late 2010s before levelling off , reflecting Netflix's massive investment in original content during that period.

# Country Analysis

## Top Producing Countries

```{r top-countries}
# Explode countries and count
country_counts <- Titles %>%
    explode_countries() %>%
    count_top_n(country, n = 15)

ggplot(country_counts, aes(x = reorder(country, n), y = n)) +
    geom_col(fill = "#E50914") +
    coord_flip() +
    labs(title = "Top 15 Content-Producing Countries on Netflix",
         x = NULL, y = "Number of Titles") +
    theme_minimal()
```

- The **US** dominates by a massive margin, followed by **India** and the **UK**. This is expected given Netflix's origins and the size of Hollywood and Bollywood.

## Genre Preferences by Country (Top 5 Countries)

```{r genre-by-country, fig.height=8}
top5_countries <- country_counts %>% slice_head(n = 5) %>% pull(country)

Titles %>%
    explode_countries() %>%
    filter(country %in% top5_countries) %>%
    explode_genres() %>%
    count(country, genre, sort = TRUE) %>%
    group_by(country) %>%
    slice_head(n = 5) %>%
    ungroup() %>%
    ggplot(aes(x = reorder_within(genre, n, country), y = n, fill = country)) +
    geom_col(show.legend = FALSE) +
    facet_wrap(~country, scales = "free_y", ncol = 1) +
    coord_flip() +
    scale_x_reordered() +
    labs(title = "Top 5 Genres in Each of the Top 5 Producing Countries",
         x = NULL, y = "Count") +
    theme_minimal() +
    scale_fill_brewer(palette = "Set1")
```

- **Drama** and **comedy** consistently appear across all major countries. **India** and **Great Britain** shows a strong preference for drama and comedy, while **Japan** leans towards animation and drama.

# Genre Analysis

## Most Popular Genres 

```{r genre-overall}
genre_counts <- Titles %>%
    explode_genres() %>%
    count_top_n(genre, n = 12)

ggplot(genre_counts, aes(x = reorder(genre, n), y = n)) +
    geom_col(fill = "#831010") +
    geom_text(aes(label = n), hjust = -0.15, size = 3.5) +
    coord_flip() +
    labs(title = "Most Common Genres on Netflix",
         x = NULL, y = "Number of Titles") +
    theme_minimal()
```

- **Drama** is by far the most common genre, followed by **comedy** and **action**. This suggests investors should prioritise drama content when building a catalogue.

## Genre Trends Over Time

```{r genre-trend, fig.height=7}
top6_genres <- genre_counts %>% slice_head(n = 6) %>% pull(genre)

Titles %>%
    explode_genres() %>%
    filter(genre %in% top6_genres, release_year >= 2000) %>%
    count(release_year, genre) %>%
    ggplot(aes(x = release_year, y = n, colour = genre)) +
    geom_line(linewidth = 1) +
    labs(title = "Genre Trends Over Time (Top 6 Genres, post-2000)",
         x = "Release Year", y = "Count", colour = "Genre") +
    theme_minimal() +
    scale_colour_brewer(palette = "Set2")
```

- All genres see exponential growth from \~2015 onwards. **Drama** consistently leads. **Documentation** (documentaries) has also seen a strong rise, suggesting growing viewer appetite for non-fiction content.

# Ratings & Scores

## IMDb Score Distribution

```{r imdb-dist}
Titles %>%
    filter(!is.na(imdb_score)) %>%
    ggplot(aes(x = imdb_score)) +
    geom_histogram(binwidth = 0.3, fill = "#E50914", colour = "white", alpha = 0.85) +
    geom_vline(aes(xintercept = median(imdb_score, na.rm = TRUE)),
               linetype = "dashed", colour = "black", linewidth = 0.8) +
    labs(title = "Distribution of IMDb Scores Across Netflix Titles",
         subtitle = "Dashed line = median score",
         x = "IMDb Score", y = "Count") +
    theme_minimal()
```

- Scores are roughly normally distributed and centre around 6-7, with relatively few titles scoring above 8. The platform carries a mix of quality and not everything is a hit.

## IMDb Scores by Genre

```{r score-by-genre}
Titles %>%
    explode_genres() %>%
    filter(genre %in% top6_genres, !is.na(imdb_score)) %>%
    ggplot(aes(x = reorder(genre, imdb_score, FUN = median), y = imdb_score, fill = genre)) +
    geom_boxplot(show.legend = FALSE, alpha = 0.8) +
    coord_flip() +
    labs(title = "IMDb Score Distribution by Genre (Top 6)",
         x = NULL, y = "IMDb Score") +
    theme_minimal() +
    scale_fill_brewer(palette = "Set2")
```

- **Documentation** (documentaries) tend to score slightly higher on average, while **action** and **comedy** show wider variance. This suggests documentaries are a safer bet quality-wise.

## Scores by Country (Top 10)

```{r score-by-country, fig.height=7}
top10_countries <- country_counts %>% slice_head(n = 10) %>% pull(country)

Titles %>%
    explode_countries() %>%
    filter(country %in% top10_countries, !is.na(imdb_score)) %>%
    ggplot(aes(x = reorder(country, imdb_score, FUN = median), y = imdb_score, fill = country)) +
    geom_boxplot(show.legend = FALSE, alpha = 0.8) +
    coord_flip() +
    labs(title = "IMDb Score Distribution by Country (Top 10 Producing Countries)",
         x = NULL, y = "IMDb Score") +
    theme_minimal()
```

- Countries like the **Korea Republic** tend to produce higher-rated content on average, while **India's** larger volume comes with a wider spread of quality.

## Age Certification Breakdown

```{r age-cert}
Titles %>%
    filter(!is.na(age_certification), age_certification != "") %>%
    count_top_n(age_certification, n = 10) %>%
    ggplot(aes(x = reorder(age_certification, n), y = n)) +
    geom_col(fill = "#564d4d") +
    coord_flip() +
    labs(title = "Content by Age Certification",
         x = NULL, y = "Count") +
    theme_minimal()
```

- A large portion of content is rated **R** or **TV-MA**, indicating Netflix skews towards adult-oriented content. Investors wanting a family-friendly platform would need to fill a different niche.

# Movie Length Analysis

## Distribution of Movie Runtimes

```{r runtime-dist}
# Filter to movies only and remove extreme outliers
movies_only <- Titles %>%
    filter(type == "MOVIE", !is.na(runtime), runtime > 0, runtime < 300)

ggplot(movies_only, aes(x = runtime)) +
    geom_histogram(binwidth = 5, fill = "#E50914", colour = "white", alpha = 0.85) +
    geom_vline(aes(xintercept = median(runtime)),
               linetype = "dashed", colour = "black", linewidth = 0.8) +
    labs(title = "Distribution of Movie Runtimes on Netflix",
         subtitle = "Dashed line = median",
         x = "Runtime (minutes)", y = "Count") +
    theme_minimal()
```

- The typical Netflix movie is around **90-100 minutes** long. There's a nice bell-shaped distribution with a slight right tail from longer epic-style movies.

## Run-time by Genre

```{r runtime-by-genre}
Titles %>%
    filter(type == "MOVIE", !is.na(runtime), runtime > 0, runtime < 300) %>%
    explode_genres() %>%
    filter(genre %in% top6_genres) %>%
    ggplot(aes(x = reorder(genre, runtime, FUN = median), y = runtime, fill = genre)) +
    geom_boxplot(show.legend = FALSE, alpha = 0.8) +
    coord_flip() +
    labs(title = "Movie Runtime by Genre",
         x = NULL, y = "Runtime (minutes)") +
    theme_minimal() +
    scale_fill_brewer(palette = "Set2")
```

- **Drama** and **action** movies tend to be the longest. **Comedy** and **animation** typically run shorter. This has cost implications, longer movies mean more production hours.

## Average Run-time Over Time

```{r runtime-trend}
movies_only %>%
    filter(release_year >= 1980) %>%
    group_by(release_year) %>%
    summarise(avg_runtime = mean(runtime, na.rm = TRUE), .groups = "drop") %>%
    ggplot(aes(x = release_year, y = avg_runtime)) +
    geom_point(alpha = 0.5, colour = "#E50914") +
    geom_smooth(method = "loess", se = TRUE, colour = "black", linewidth = 0.8) +
    labs(title = "Average Movie Runtime Over Time",
         x = "Release Year", y = "Average Runtime (min)") +
    theme_minimal()
```

- There's a slight **decline** in average run-time in recent years, suggesting a trend towards shorter, more digestible content — likely influenced by streaming habits where viewers have shorter attention spans.

# Netflix Movies Deep Dive

## Duration Distribution 

```{r movieinfo-duration}
Movie_Info %>%
    filter(!is.na(duration_mins), duration_mins > 0, duration_mins < 300) %>%
    ggplot(aes(x = duration_mins)) +
    geom_histogram(binwidth = 5, fill = "#831010", colour = "white", alpha = 0.85) +
    labs(title = "Duration of Netflix Movies (Movie_Info Dataset)",
         x = "Duration (minutes)", y = "Count") +
    theme_minimal()
```

## Top Movie Categories 

```{r top-categories}
Movie_Info %>%
    explode_listed_in() %>%
    count_top_n(category, n = 12) %>%
    ggplot(aes(x = reorder(category, n), y = n)) +
    geom_col(fill = "#E50914") +
    coord_flip() +
    labs(title = "Most Common Movie Categories on Netflix",
         subtitle = "From the listed_in column in Movie_Info",
         x = NULL, y = "Count") +
    theme_minimal()
```

- **International Movies**, **Dramas**, and **Comedies** lead the pack. Netflix clearly invests heavily in international content to serve its global audience.

## Top Countries 

```{r movieinfo-countries}
# Split the country column (comma-separated)
Movie_Info %>%
    separate_rows(country, sep = ",\\s*") %>%
    mutate(country = str_trim(country)) %>%
    filter(!is.na(country), country != "") %>%
    count_top_n(country, n = 10) %>%
    ggplot(aes(x = reorder(country, n), y = n)) +
    geom_col(fill = "#564d4d") +
    coord_flip() +
    labs(title = "Top 10 Countries Producing Netflix Movies",
         x = NULL, y = "Count") +
    theme_minimal()
```

# Textual Analysis of Descriptions

We now look at the most common words used in Netflix movie descriptions to see what themes and topics dominate the platform.

## Word Frequencies (Titles Descriptions)

```{r text-analysis}
# Use our functional text pipeline from the helper functions
word_freq <- Titles %>%
    filter(type == "MOVIE") %>%
    pull(description) %>%
    get_word_freq(n = 25)

ggplot(word_freq, aes(x = reorder(word, n), y = n)) +
    geom_col(fill = "#E50914") +
    coord_flip() +
    labs(title = "Most Common Words in Netflix Movie Descriptions",
         subtitle = "Stop words removed",
         x = NULL, y = "Frequency") +
    theme_minimal()
```

- Words like **"life"**, **"family"**, **"love"**, **"world"** and **"young"** dominate in titles. This tells us Netflix movies heavily revolve around relatable, human-centric themes and stories about people navigating life, relationships and coming-of-age.

## Word Frequencies by Type (Movie vs Show)

```{r text-by-type, fig.height=8}
# Use map to get word frequencies for each type
word_by_type <- Titles %>%
    split(.$type) %>%
    map(~.x %>% pull(description) %>% get_word_freq(n = 15)) %>%
    imap_dfr(~.x %>% mutate(type = .y))

ggplot(word_by_type, aes(x = reorder_within(word, n, type), y = n, fill = type)) +
    geom_col(show.legend = FALSE) +
    facet_wrap(~type, scales = "free") +
    coord_flip() +
    scale_x_reordered() +
    labs(title = "Top Words in Descriptions: Movies vs Shows",
         x = NULL, y = "Frequency") +
    theme_minimal() +
    scale_fill_manual(values = c("MOVIE" = "#E50914", "SHOW" = "#564d4d"))
```

- Movies tend to focus more on individual stories ("life", "young", "love"), while shows lean towards group dynamics and ongoing narratives ("group", "series", "team").

# Notable Actors & Directors

## Most Prolific Actors on Netflix

```{r top-actors}
# Filter to actors only and count appearances
top_actors <- Credits %>%
    filter(role == "ACTOR") %>%
    count_top_n(name, n = 15)

ggplot(top_actors, aes(x = reorder(name, n), y = n)) +
    geom_col(fill = "#E50914") +
    coord_flip() +
    labs(title = "Top 15 Most Prolific Actors on Netflix",
         x = NULL, y = "Number of Titles") +
    theme_minimal()
```

## Most Prolific Directors

```{r top-directors}
top_directors <- Credits %>%
    filter(role == "DIRECTOR") %>%
    count_top_n(name, n = 15)

ggplot(top_directors, aes(x = reorder(name, n), y = n)) +
    geom_col(fill = "#831010") +
    coord_flip() +
    labs(title = "Top 15 Most Prolific Directors on Netflix",
         x = NULL, y = "Number of Titles") +
    theme_minimal()
```

# Key Takeaways

Based on the analysis, here are the main insights for investors looking to launch a competing streaming platform:

1.  **Content Mix**: Netflix is **movie-heavy**. A new platform could differentiate by investing more in original series, which build subscriber loyalty through ongoing engagement.

2.  **Geography**: The **US, India, and UK** dominate production. There's room to tap into underrepresented markets (Africa, Southeast Asia, Latin America) where content demand is growing.

3.  **Genre Strategy**: **Drama** is king, but **documentaries** consistently score well and are cheaper to produce. Comedy and action have broad appeal but variable quality.

4.  **Quality**: The median IMDb score hovers around 6.5 , a lot of Netflix content is "average." A quality-over-quantity strategy could be a competitive edge.

5.  **Movie Length**: Movies are trending **shorter**, reflecting streaming viewing habits. Content under 100 minutes tends to perform well. This can also be due to the recent rise in Tiktok and Instagram reels.

6.  **Themes**: Descriptions revolve around **life, family, love, and coming-of-age** stories. These universal themes have broad appeal across demographics.

7.  **Talent**: A few actors and directors appear repeatedly securing exclusive deals with prolific talent could drive viewership.
