---
title: "Question 3: Loans and Credit"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.width = 9, fig.height = 5.5)

if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, lubridate, janitor, scales, broom,
               knitr, patchwork, glue)

list.files('code/', full.names = T, recursive = T) %>%
    .[grepl('.R', .)] %>%
    as.list() %>%
    walk(~source(.))

```

# 1. Loading and Cleaning

```{r load}
Loan_raw <- load_loan_data("data/Loan_Cred/loan_data.rds")
glue("Raw data: {nrow(Loan_raw)} loans across {ncol(Loan_raw)} columns")
```

The dataset has a lot of columns, many of which are mostly empty. We clean by removing those, fixing formatting issues, and creating a clear loan outcome label.

```{r clean}
Loan_df <- Loan_raw %>%
    clean_names() %>%
    {if ("int_rate"   %in% names(.)) mutate(., int_rate   = clean_int_rate(int_rate))   else .} %>%
    {if ("emp_length" %in% names(.)) mutate(., emp_length = clean_emp_length(emp_length)) else .} %>%
    {if ("term"       %in% names(.)) mutate(., term       = clean_term(term))             else .} %>%
    drop_mostly_empty(cutoff = 0.80) %>%
    mutate(
        loan_outcome = classify_loan(loan_status),
        is_default   = as.integer(loan_outcome == "Default")
    )

glue("After cleaning: {nrow(Loan_df)} loans, {ncol(Loan_df)} columns")
```

```{r add-features}
# find whichever FICO columns survived
fico_low  <- names(Loan_df)[str_detect(names(Loan_df), "fico.*low")][1]
fico_high <- names(Loan_df)[str_detect(names(Loan_df), "fico.*high")][1]

Loan_df <- Loan_df %>%
    mutate(
        fico_mid = if (!is.na(fico_low) & !is.na(fico_high))
            (.data[[fico_low]] + .data[[fico_high]]) / 2 else NA_real_,

        income_band = cut(annual_inc,
                          breaks = c(0, 30000, 50000, 75000, 100000, 150000, Inf),
                          labels = c("<30k","30-50k","50-75k","75-100k","100-150k","150k+")),

        dti_band = cut(dti, breaks = c(0, 10, 15, 20, 25, 30, Inf),
                       labels = c("<10","10-15","15-20","20-25","25-30","30+")),

        emp_group = case_when(
            is.na(emp_length) ~ "Unknown",
            emp_length == 0   ~ "<1 yr",
            emp_length <= 3   ~ "1-3 yrs",
            emp_length <= 5   ~ "4-5 yrs",
            emp_length <= 9   ~ "6-9 yrs",
            TRUE              ~ "10+ yrs"),

        term_label = if_else(term == 36, "36 months", "60 months")
    )
```

How the loans break down by outcome:

```{r outcome}
Loan_df %>%
    count(loan_outcome, sort = TRUE) %>%
    mutate(share = sprintf("%.1f%%", n / sum(n) * 100)) %>%
    kable(col.names = c("Outcome", "Count", "Share"), format = "pipe")
```

**Reading the outcome split.** We start with **1,000,000 loans and 145 columns**; dropping any column that is more than 80% empty leaves 108 usable columns without throwing away a single loan. Most of the book is still **live** — 59.8% of loans are *Current* and haven't reached a final outcome yet, so they can't tell us anything about repayment. *Fully Paid* (29.7%) and *Default* (8.6%) are the only **resolved** outcomes, with a small *Other* bucket (1.9%, e.g. charged-off/grace categories that don't map cleanly).

From here we focus on **closed loans only** (Fully Paid vs Default) so the comparisons are fair.

```{r closed}
closed <- Loan_df %>% filter(loan_outcome %in% c("Fully Paid", "Default"))
glue("{nrow(closed)} closed loans — overall default rate: {sprintf('%.1f%%', mean(closed$is_default)*100)}")
```

**Why the default rate jumps to 22.5%.** Restricting to the **383,202 resolved loans** changes the denominator: defaults are 8.6% of the *whole* book but **22.5% of loans that have actually finished** (86,079 ÷ 383,202). This conditional rate is the honest baseline for the rest of the analysis — every group default rate below should be read against this **22.5% benchmark**: above it = riskier than average, below it = safer.

---

# 2. What Drives Defaults?

We check default rates across the main borrower characteristics using `map` to loop through them all at once.

```{r rates}
vars_to_check <- c("grade", "purpose", "term_label",
                    "home_ownership", "emp_group", "income_band", "dti_band")

rate_tables <- vars_to_check %>%
    set_names() %>%
    map(~ default_rate_by(Loan_df, .x))
```

## Credit Grade

```{r grade-plot, fig.height=4}
rate_tables$grade %>%
    ggplot(aes(x = grade, y = default_rate, fill = default_rate)) +
    geom_col(show.legend = FALSE) +
    geom_text(aes(label = sprintf("%.1f%%", default_rate*100)), vjust = -0.5, size = 3.5) +
    scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.12))) +
    scale_fill_gradient(low = "#2C7A4B", high = "#C0392B") +
    labs(title = "Default rate by credit grade",
         subtitle = "Grade A borrowers are safest; risk climbs steadily toward G",
         x = "Grade", y = "Default Rate") +
    theme_q3()
```

**Credit grade is the single strongest signal.** Default rates climb almost perfectly monotonically — **A 6.9% → B 15.6% → C 25.3% → D 34.6% → E 43.1% → F 52.4% → G 57.3%**. A G-grade borrower is **roughly 8× more likely to default than an A-grade borrower**, and by grade E it is closer to a coin-flip than a loan. The grade boundary that matters most operationally is **C (25.3%), which is already above the 22.5% book average** — so risk crosses the "worse than typical" line as early as grade C, not at the D–G tail.

## Loan Purpose

```{r purpose-plot, fig.height=5}
rate_tables$purpose %>%
    filter(n_loans >= 100) %>%
    ggplot(aes(x = reorder(purpose, default_rate), y = default_rate)) +
    geom_col(fill = "#C0392B", alpha = 0.8) +
    geom_text(aes(label = sprintf("%.1f%%", default_rate*100)), hjust = -0.1, size = 3) +
    scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.15))) +
    coord_flip() +
    labs(title = "Default rate by loan purpose",
         subtitle = "Small business loans are the riskiest by far",
         x = NULL, y = "Default Rate") +
    theme_q3()
```

**Small-business lending stands out as the riskiest purpose.** At **33.3%** it is well clear of the next tier — *renewable energy* (25.7%), *moving* (25.2%) and *medical* (24.7%) — all of which sit just above the book average. The largest category in practice, **debt consolidation (23.8%)**, is essentially average. The safest reasons to borrow are asset-backed or routine: **car (16.6%)**, **home improvement (19.3%)** and **credit card (19.5%)** all sit comfortably below 22.5%. The spread (16.6% → 33.3%) is meaningful but much narrower than the grade spread, so purpose is a useful secondary filter rather than a primary one.

## Loan Term and Home Ownership

```{r term-home, fig.height=4}
p1 <- rate_tables$term_label %>%
    ggplot(aes(x = term_label, y = default_rate, fill = term_label)) +
    geom_col(show.legend = FALSE) +
    geom_text(aes(label = sprintf("%.1f%%", default_rate*100)), vjust = -0.5, size = 3.5) +
    scale_y_continuous(labels = percent, expand = expansion(mult = c(0,0.15))) +
    scale_fill_manual(values = c("#2980B9","#E74C3C")) +
    labs(title = "By loan term", x = NULL, y = "Default Rate") + theme_q3()

p2 <- rate_tables$home_ownership %>%
    filter(n_loans >= 100) %>%
    ggplot(aes(x = reorder(home_ownership, default_rate), y = default_rate)) +
    geom_col(fill = "#8E44AD", alpha = 0.75) +
    geom_text(aes(label = sprintf("%.1f%%", default_rate*100)), vjust = -0.5, size = 3.5) +
    scale_y_continuous(labels = percent, expand = expansion(mult = c(0,0.15))) +
    labs(title = "By home ownership", x = NULL, y = NULL) + theme_q3()

p1 + p2
```

**Term length is a major driver; home ownership is a minor one.** Stretching a loan from 36 to 60 months nearly **doubles** the default rate (**19.0% → 34.5%**) — longer terms mean more time for a borrower's circumstances to deteriorate, and 60-month loans also tend to go to weaker credits. Home ownership moves the needle far less: **MORTGAGE 18.8% < OWN 22.8% < RENT 27.0%**. Renters default about 8 points more than mortgage-holders, but note the ordering — borrowers who **own outright (22.8%) default slightly more than those still paying a mortgage (18.8%)**, suggesting the mortgage group captures vetted, higher-income borrowers rather than ownership *per se* being protective.

## Income and DTI

```{r income-dti, fig.height=4}
p3 <- rate_tables$income_band %>%
    filter(n_loans >= 50) %>%
    ggplot(aes(x = income_band, y = default_rate)) +
    geom_col(fill = "#27AE60", alpha = 0.8) +
    geom_text(aes(label = sprintf("%.1f%%", default_rate*100)), vjust = -0.5, size = 3) +
    scale_y_continuous(labels = percent, expand = expansion(mult = c(0,0.12))) +
    labs(title = "By income band", x = NULL, y = "Default Rate") + theme_q3()

p4 <- rate_tables$dti_band %>%
    filter(n_loans >= 50) %>%
    ggplot(aes(x = dti_band, y = default_rate)) +
    geom_col(fill = "#E67E22", alpha = 0.8) +
    geom_text(aes(label = sprintf("%.1f%%", default_rate*100)), vjust = -0.5, size = 3) +
    scale_y_continuous(labels = percent, expand = expansion(mult = c(0,0.12))) +
    labs(title = "By DTI band", x = NULL, y = NULL) + theme_q3()

p3 + p4
```

**Both income and DTI behave exactly as theory predicts, and both are smooth gradients rather than cliffs.** Default risk falls steadily with income — **<30k 28.7% → 150k+ 16.1%** — a ~13-point swing from the poorest to the richest band. It rises just as steadily with leverage — **DTI <10 16.8% → 30+ 31.7%** — roughly doubling across the range. Two practical notes: the **`NA` bars (income 20.1%, DTI 21.5%) sit near the average**, so missing values aren't hiding unusually risky borrowers; and because the DTI relationship is a *gradient with no sharp break*, any hard DTI cut-off (Section 4) is a judgement call about where on the slope to draw the line, not a natural cliff in the data.

---

# 3. Who Defaults Most?

Putting it all together, the riskiest borrower profile looks like this:

```{r risky-profile}
tibble(
    Factor = c("Credit Grade", "Loan Purpose", "Loan Term",
               "Income", "DTI", "Home Ownership"),
    `Highest Default Risk` = c(
        "Grade F / G",
        "Small business loans",
        "60-month (long term)",
        "Below $30k",
        "Above 25",
        "Renters (slightly higher)")
) %>%
    kable(format = "pipe")
```

**The risk factors stack.** None of these is the whole story on its own, but they compound: a low-grade, low-income borrower taking a 60-month small-business loan at high DTI carries risk from every column at once, which is exactly the profile that pushes well past the 22.5% baseline toward the 40–55% range seen in the grade and interest-rate views.

And if we look at interest rate — borrowers paying higher rates are more likely to default:

```{r int-rate-plot, fig.height=4}
closed %>%
    mutate(rate_band = cut(int_rate, breaks = seq(4, 28, by = 3), include.lowest = TRUE)) %>%
    filter(!is.na(rate_band)) %>%
    group_by(rate_band) %>%
    summarise(default_rate = mean(is_default), n = n(), .groups = "drop") %>%
    filter(n >= 30) %>%
    ggplot(aes(x = rate_band, y = default_rate, group = 1)) +
    geom_line(linewidth = 1, color = "#2980B9") +
    geom_point(size = 2.5, color = "#2980B9") +
    scale_y_continuous(labels = percent) +
    labs(title = "Default rate by interest rate",
         subtitle = "Higher rates correlate with more defaults",
         x = "Interest Rate Range (%)", y = "Default Rate") +
    theme_q3() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Interest rate is the cleanest monotonic predictor of all.** Default rate climbs almost in a straight line from **~5.5% in the 4–7% rate band to ~49% in the 25–28% band** — close to a 9× gradient with no reversal. This is partly mechanical (the lender already prices observed risk into the rate, so rate is a *summary* of grade, term and DTI), and partly a feedback loop worth flagging: the highest-rate borrowers are the ones least able to absorb a high rate, so the pricing that is meant to compensate for risk may itself **add** to it. Because rate bundles the other signals, it is the most useful single field for a quick risk read, but it should not be treated as an *independent* cause.

---

# 4. DTI Hard-Cap: What Should It Be?

The Director wants a recommended Debt-to-Income (DTI) cutoff. Let's look at how default rate changes as DTI rises.

```{r dti-curve, fig.height=4.5}
dti_curve <- closed %>%
    filter(!is.na(dti)) %>%
    mutate(dti_bin = cut(dti, breaks = seq(0, 40, by = 2), include.lowest = TRUE)) %>%
    group_by(dti_bin) %>%
    summarise(default_rate = mean(is_default), n = n(), .groups = "drop") %>%
    filter(n >= 30)

dti_curve %>%
    ggplot(aes(x = dti_bin, y = default_rate, group = 1)) +
    geom_line(linewidth = 1, color = "#2980B9") +
    geom_point(size = 2.5, color = "#2980B9") +
    scale_y_continuous(labels = percent) +
    labs(title = "Default rate by DTI",
         subtitle = "Risk picks up noticeably above DTI of ~25",
         x = "DTI Range", y = "Default Rate") +
    theme_q3() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**The curve has a flat floor and then a steady climb — no clean cliff.** From DTI 0 to ~10 the default rate is essentially flat at its **floor of ~17%** (it even dips slightly around 8–10). It then rises continuously: ~20% by DTI 14, ~24% by DTI 22, ~28% by DTI 28, and into the **32–34% range above DTI 34**. The key takeaway for setting a cap is that the curve **accelerates through the low-to-mid 20s** — that is the "elbow" the subtitle refers to — but there is no single DTI value where risk suddenly explodes, so the cap is a risk-appetite choice along a smooth slope.

Suggested caps depending on how much risk the Institute is willing to accept:

```{r dti-caps}
tolerances <- c(0.12, 0.15, 0.18, 0.20)

dti_caps <- map_dfr(tolerances, function(tol) {
    cap <- dti_curve %>%
        filter(default_rate <= tol) %>%
        pull(dti_bin) %>%
        as.character() %>%
        str_extract("\\d+(?=\\])") %>%
        as.numeric() %>%
        max(na.rm = TRUE)
    tibble(`Risk tolerance` = sprintf("%.0f%%", tol*100),
           `Suggested DTI cap` = cap)
})

dti_caps %>% kable(format = "pipe")
```

**Important: read this table together with the floor of the curve.** The rule here is *"find the highest DTI band whose default rate still sits at or below the chosen tolerance."* Because the **minimum default rate anywhere on the curve is ~16.8%**, any tolerance *below* that floor is unachievable — that is why **12% and 15% return `-Inf`** (no band qualifies, so the maximum of an empty set is `-Inf`). Only the **18% (cap ≈ 12)** and **20% (cap ≈ 16)** tolerances are attainable, and even those force a strict cut-off because the whole closed book averages 22.5%.

So there are **two legitimate ways to set the cap, and they give different numbers**:

- **Absolute-tolerance rule (this table):** to hold defaults at/under ~18–20%, the cap must be tight — **DTI ≈ 12–16**.
- **Elbow / "where risk accelerates" rule (the curve):** if the goal is instead to cut off the steep part of the slope rather than hit an absolute rate, the cap lands higher, around **DTI 20–25**.

The right choice depends on the Director's objective (a hard loss target vs. trimming the worst of the tail); the recommendation in Section 7 is framed around the elbow, but the absolute-tolerance numbers above are the stricter, loss-target answer.

## Does it differ by state?

```{r dti-state, fig.height=4.5}
closed %>%
    filter(!is.na(dti)) %>%
    mutate(region = if_else(addr_state == "TX", "Texas", "Other States"),
           dti_bin = cut(dti, breaks = seq(0, 40, by = 3), include.lowest = TRUE)) %>%
    group_by(dti_bin, region) %>%
    summarise(default_rate = mean(is_default), n = n(), .groups = "drop") %>%
    filter(n >= 20) %>%
    ggplot(aes(x = dti_bin, y = default_rate, color = region, group = region)) +
    geom_line(linewidth = 1) + geom_point(size = 2) +
    scale_color_manual(values = c("Texas" = "#C0392B", "Other States" = "#2980B9")) +
    scale_y_continuous(labels = percent) +
    labs(title = "DTI vs Default: Texas compared to the rest",
         x = "DTI Range", y = "Default Rate", color = NULL) +
    theme_q3() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Texas and the rest of the country trace essentially the same DTI curve.** The two lines overlap closely across the whole range — Texas is marginally higher at very low DTI and marginally lower in the low-20s, but the differences are within noise and both rise from ~17% to the low-30s in lockstep. There is **no evidence that the DTI–default relationship is structurally different in Texas**, which is what justifies a single national cap rather than a Texas-specific one.

---

# 5. Is Texas Different?

```{r texas-table}
closed %>%
    mutate(region = if_else(addr_state == "TX", "Texas", "Other States")) %>%
    group_by(region) %>%
    summarise(
        Loans = n(),
        `Default Rate` = sprintf("%.1f%%", mean(is_default)*100),
        `Avg Loan` = sprintf("$%s", comma(round(mean(loan_amnt)))),
        `Avg Income` = sprintf("$%s", comma(round(mean(annual_inc)))),
        `Avg DTI` = round(mean(dti, na.rm = TRUE), 1),
        `Avg Interest Rate` = sprintf("%.1f%%", mean(int_rate, na.rm = TRUE)),
        .groups = "drop"
    ) %>%
    kable(format = "pipe")
```

**On the headline numbers Texas is unremarkable.** Its **default rate (22.8%) is just 0.4 points above the rest of the country (22.4%)**, on a large sample of 32,383 Texan loans. The profile differences are small and mostly cancel out: Texan borrowers carry slightly **higher income ($83.5k vs $78.4k)** *and* slightly **higher leverage (DTI 19.8 vs 18.7)** and larger loans, while being charged an **identical average interest rate (13.0%)**. In other words, Texas looks like a slightly higher-income, slightly more-leveraged version of the national book, and the two effects roughly offset.

```{r texas-test}
tx <- closed %>% mutate(is_tx = addr_state == "TX")
ptest <- prop.test(
    x = c(sum(tx$is_default[tx$is_tx]), sum(tx$is_default[!tx$is_tx])),
    n = c(sum(tx$is_tx), sum(!tx$is_tx))
)
```

Proportion test p-value: **`r format.pval(ptest$p.value, digits = 3)`**. Texas is **`r if_else(ptest$p.value < 0.05, "statistically different", "not significantly different")`** from the national average.

**Interpreting the test.** With a p-value of **0.151 (> 0.05)** we cannot reject the null that Texas defaults at the same rate as everywhere else. Note this is a case where the **huge sample makes the test sensitive to even tiny gaps**, yet the 0.4-point difference is still *not* significant — strong evidence that the gap is genuinely just noise rather than a real Texas effect.

## How do all states compare?

```{r state-plot, fig.height=7}
closed %>%
    group_by(addr_state) %>%
    summarise(n = n(), default_rate = mean(is_default), .groups = "drop") %>%
    filter(n >= 200) %>%
    mutate(is_tx = addr_state == "TX") %>%
    ggplot(aes(x = reorder(addr_state, default_rate), y = default_rate)) +
    geom_segment(aes(xend = reorder(addr_state, default_rate),
                     y = 0, yend = default_rate), color = "grey70") +
    geom_point(aes(color = is_tx, size = n)) +
    scale_color_manual(values = c("grey40", "#C0392B"),
                       labels = c("Other", "Texas"), name = NULL) +
    scale_size_continuous(name = "Loans", labels = comma) +
    scale_y_continuous(labels = percent) +
    coord_flip() +
    labs(title = "Default rates by state",
         subtitle = "Texas in red — states with 200+ loans shown",
         x = NULL, y = "Default Rate") +
    theme_q3()
```

**Texas sits squarely in the middle of the pack.** The state spread runs from roughly **~13% (DC, ME, OR at the safe end) to ~26–27% (AR, LA, MS, OK, AL at the risky end)** — a real ~13-point range, so states genuinely do differ overall (see Belief 2). But **Texas (red) lands mid-table at ~22.8%**, close to the national average and nowhere near either extreme. The riskiest states cluster in the Deep South, while the safest are a mix of small-volume states; Texas is simply a big, average market.

---

# 6. Testing the Institute's Beliefs

The Institute holds three long-standing beliefs. Let's check them against the data.

## Belief 1: Homeowners employed 10+ years default less on short-term loans

```{r h1}
closed %>%
    filter(term == 36) %>%
    mutate(group = case_when(
        home_ownership %in% c("OWN","MORTGAGE") & emp_length >= 10 ~ "Homeowner & 10+ yrs",
        TRUE ~ "Everyone else"
    )) %>%
    group_by(group) %>%
    summarise(Loans = n(),
              `Default Rate` = sprintf("%.1f%%", mean(is_default)*100),
              .groups = "drop") %>%
    kable(format = "pipe")
```

**Directionally true, but the effect is moderate.** Among 36-month loans, the "homeowner **and** 10+ years employed" group defaults at **14.6%** versus **20.3%** for everyone else — a **5.7-point** improvement, or about **28% lower** in relative terms. That is a real, useful reduction, but the favoured group still defaults at nearly **1 in 7**, so the combination is a helpful *positive signal*, not a guarantee of repayment. Hence **partially supported** — stability helps, but it doesn't override grade, term or DTI risk.

## Belief 2: States differ significantly on short-term defaults

```{r h2}
h2_data <- closed %>% filter(term == 36)
h2_test <- chisq.test(table(h2_data$addr_state, h2_data$is_default),
                       simulate.p.value = TRUE)
```

Chi-squared p-value: **`r format.pval(h2_test$p.value, digits = 4)`** — this is **`r if_else(h2_test$p.value < 0.05, "significant, so yes, states do differ. Supported.", "not significant. Not supported.")`**

**Interpreting the test.** The chi-squared test (with simulated p-values, appropriate given the many sparse state cells) returns **p ≈ 0.0005**, far below 0.05, so the differences in short-term default rates across states are **statistically real, not random** — consistent with the ~13-point spread seen in the state plot. **Supported.** One caveat to keep in mind: "states differ" does **not** mean the *state itself* causes default — much of the gap likely reflects different borrower mixes (income, grade, purpose) across states rather than geography.

## Belief 3: Credit grades predict defaults well for younger borrowers

We look at borrowers with lower FICO scores (below 680) as a proxy for less established credit histories.

```{r h3, fig.height=4}
has_fico <- "fico_mid" %in% names(closed) && !all(is.na(closed$fico_mid))

if (has_fico) {
    h3 <- closed %>% filter(fico_mid < 680)
    h3_sub <- "Borrowers with FICO < 680"
} else {
    h3 <- closed
    h3_sub <- "All borrowers (FICO not available)"
}

h3 %>%
    group_by(grade) %>%
    summarise(default_rate = mean(is_default), n = n(), .groups = "drop") %>%
    ggplot(aes(x = grade, y = default_rate, fill = grade)) +
    geom_col(show.legend = FALSE) +
    geom_text(aes(label = sprintf("%.1f%%", default_rate*100)), vjust = -0.5, size = 3.5) +
    scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.12))) +
    scale_fill_brewer(palette = "YlOrRd") +
    labs(title = "Default rates by grade", subtitle = h3_sub,
         x = "Grade", y = "Default Rate") +
    theme_q3()
```

**Grade clearly separates risk — with one honest caveat about the test.** The step-up from **A 6.9% to G 57.3%** is exactly what a well-functioning grading system should produce, so grade is a strong predictor. **However**, the FICO columns did not survive cleaning (`fico_mid` is all `NA`), so the code fell back to its `else` branch and the subtitle reads *"All borrowers (FICO not available)"*. That means this chart shows grade's predictive power **across the whole closed book, not specifically the low-FICO / thin-file "younger" borrowers** the belief is about. The belief is therefore **supported for borrowers in general**, but with the data available we **cannot isolate the younger-borrower sub-group** — to test the belief as literally stated we would need the FICO fields restored upstream.

## Summary of beliefs

```{r h-summary}
tibble(
    Belief = c(
        "Homeowners + 10yr employment default less (short-term)",
        "States differ on short-term defaults",
        "Credit grades predict default for younger borrowers"
    ),
    Verdict = c(
        "Partially supported — small difference",
        if_else(h2_test$p.value < 0.05, "Supported", "Not supported"),
        "Supported — clear pattern across grades"
    )
) %>% kable(format = "pipe")
```

**In short:** one belief is fully borne out (states differ), one holds at the overall level but couldn't be tested on the exact sub-group it claims (grades predict for younger borrowers — FICO missing), and one is real but smaller than the Institute assumes (homeowner + tenure).

---

# 7. Recommendations for the Director

Based on the analysis, here are the main takeaways:

**On DTI caps:**
The DTI–default relationship is a smooth slope, not a cliff, so the cap is a risk-appetite decision. If the goal is to trim the steep part of the curve, a cap in the **DTI 20–25** range is reasonable (risk accelerates through the low-20s). If instead the goal is an absolute loss target, the numbers are stricter: holding defaults at/under ~18–20% requires a cap of only **DTI 12–16**, because the curve's floor is already ~17%. Either way a **single national cap** is justified — Texas and the rest of the country follow the same DTI curve.

**On who defaults:**
The strongest signals are **credit grade** (6.9% at A → 57.3% at G) and **interest rate** (≈5% → ≈49%), followed by **loan term** (36mo 19.0% vs 60mo 34.5%), **purpose** (small business 33.3% is the clear outlier), and the **income/DTI gradients**. Home ownership and employment length matter, but noticeably less than grade, term and DTI.

**On credit grades:**
The grading system works well and is the best single primary filter. It is even stronger when paired with DTI and (once restored) FICO checks, especially for the borderline **C/D grades that straddle the 22.5% book average**.

**On interest rates:**
High-rate borrowers default the most. Because the rate is set to compensate for risk yet may also worsen it, it is worth reviewing whether the very highest rate bands are pricing borrowers into failure rather than out of it.

**On Texas:**
Texas is statistically in line with the national book (22.8% vs 22.4%, p = 0.151) and mid-table among states — **no special treatment or separate policy needed.**

**On the Institute's beliefs:**
Two of three hold up. States genuinely differ; credit grades clearly separate risk (though the low-FICO sub-test couldn't be run because FICO data was dropped in cleaning); and the homeowner-plus-tenure effect is real but smaller (a ~5.7-point edge) than the Institute believes. A useful follow-up would be to **restore the FICO fields upstream** so Belief 3 can be tested on the borrowers it actually concerns.
