---
title: "Question 3: Loans and Credit"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.width = 9, fig.height = 5.5)

if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, lubridate, janitor, scales, broom,
               knitr, patchwork, glue)

list.files('code/', full.names = T, recursive = T) %>%
    .[grepl('.R', .)] %>%
    as.list() %>%
    walk(~source(.))
```

# 1. Loading and Cleaning

```{r load}
Loan_raw <- load_loan_data("data/Loan_Cred/loan_data.rds")
glue("Raw data: {nrow(Loan_raw)} loans across {ncol(Loan_raw)} columns")
```

The dataset has a lot of columns, many of which are mostly empty. We clean by removing those, fixing formatting issues, and creating a clear loan outcome label.

```{r clean}
Loan_df <- Loan_raw %>%
    clean_names() %>%
    {if ("int_rate"   %in% names(.)) mutate(., int_rate   = clean_int_rate(int_rate))   else .} %>%
    {if ("emp_length" %in% names(.)) mutate(., emp_length = clean_emp_length(emp_length)) else .} %>%
    {if ("term"       %in% names(.)) mutate(., term       = clean_term(term))             else .} %>%
    drop_mostly_empty(cutoff = 0.80) %>%
    mutate(
        loan_outcome = classify_loan(loan_status),
        is_default   = as.integer(loan_outcome == "Default")
    )

glue("After cleaning: {nrow(Loan_df)} loans, {ncol(Loan_df)} columns")
```

```{r add-features}
# find whichever FICO columns survived
fico_low  <- names(Loan_df)[str_detect(names(Loan_df), "fico.*low")][1]
fico_high <- names(Loan_df)[str_detect(names(Loan_df), "fico.*high")][1]

Loan_df <- Loan_df %>%
    mutate(
        fico_mid = if (!is.na(fico_low) & !is.na(fico_high))
            (.data[[fico_low]] + .data[[fico_high]]) / 2 else NA_real_,

        income_band = cut(annual_inc,
                          breaks = c(0, 30000, 50000, 75000, 100000, 150000, Inf),
                          labels = c("<30k","30-50k","50-75k","75-100k","100-150k","150k+")),

        dti_band = cut(dti, breaks = c(0, 10, 15, 20, 25, 30, Inf),
                       labels = c("<10","10-15","15-20","20-25","25-30","30+")),

        emp_group = case_when(
            is.na(emp_length) ~ "Unknown",
            emp_length == 0   ~ "<1 yr",
            emp_length <= 3   ~ "1-3 yrs",
            emp_length <= 5   ~ "4-5 yrs",
            emp_length <= 9   ~ "6-9 yrs",
            TRUE              ~ "10+ yrs"),

        term_label = if_else(term == 36, "36 months", "60 months")
    )
```

How the loans break down by outcome:

```{r outcome}
Loan_df %>%
    count(loan_outcome, sort = TRUE) %>%
    mutate(share = sprintf("%.1f%%", n / sum(n) * 100)) %>%
    kable(col.names = c("Outcome", "Count", "Share"), format = "pipe")
```

From here we focus on **closed loans only** (Fully Paid vs Default) so the comparisons are fair.

```{r closed}
closed <- Loan_df %>% filter(loan_outcome %in% c("Fully Paid", "Default"))
glue("{nrow(closed)} closed loans — overall default rate: {sprintf('%.1f%%', mean(closed$is_default)*100)}")
```

---

# 2. What Drives Defaults?

We check default rates across the main borrower characteristics using `map` to loop through them all at once.

```{r rates}
vars_to_check <- c("grade", "purpose", "term_label",
                    "home_ownership", "emp_group", "income_band", "dti_band")

rate_tables <- vars_to_check %>%
    set_names() %>%
    map(~ default_rate_by(Loan_df, .x))
```

## Credit Grade

```{r grade-plot, fig.height=4}
rate_tables$grade %>%
    ggplot(aes(x = grade, y = default_rate, fill = default_rate)) +
    geom_col(show.legend = FALSE) +
    geom_text(aes(label = sprintf("%.1f%%", default_rate*100)), vjust = -0.5, size = 3.5) +
    scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.12))) +
    scale_fill_gradient(low = "#2C7A4B", high = "#C0392B") +
    labs(title = "Default rate by credit grade",
         subtitle = "Grade A borrowers are safest; risk climbs steadily toward G",
         x = "Grade", y = "Default Rate") +
    theme_q3()
```

Grades clearly separate good borrowers from risky ones — A-grade borrowers default far less than F or G.

## Loan Purpose

```{r purpose-plot, fig.height=5}
rate_tables$purpose %>%
    filter(n_loans >= 100) %>%
    ggplot(aes(x = reorder(purpose, default_rate), y = default_rate)) +
    geom_col(fill = "#C0392B", alpha = 0.8) +
    geom_text(aes(label = sprintf("%.1f%%", default_rate*100)), hjust = -0.1, size = 3) +
    scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.15))) +
    coord_flip() +
    labs(title = "Default rate by loan purpose",
         subtitle = "Small business loans are the riskiest by far",
         x = NULL, y = "Default Rate") +
    theme_q3()
```

## Loan Term and Home Ownership

```{r term-home, fig.height=4}
p1 <- rate_tables$term_label %>%
    ggplot(aes(x = term_label, y = default_rate, fill = term_label)) +
    geom_col(show.legend = FALSE) +
    geom_text(aes(label = sprintf("%.1f%%", default_rate*100)), vjust = -0.5, size = 3.5) +
    scale_y_continuous(labels = percent, expand = expansion(mult = c(0,0.15))) +
    scale_fill_manual(values = c("#2980B9","#E74C3C")) +
    labs(title = "By loan term", x = NULL, y = "Default Rate") + theme_q3()

p2 <- rate_tables$home_ownership %>%
    filter(n_loans >= 100) %>%
    ggplot(aes(x = reorder(home_ownership, default_rate), y = default_rate)) +
    geom_col(fill = "#8E44AD", alpha = 0.75) +
    geom_text(aes(label = sprintf("%.1f%%", default_rate*100)), vjust = -0.5, size = 3.5) +
    scale_y_continuous(labels = percent, expand = expansion(mult = c(0,0.15))) +
    labs(title = "By home ownership", x = NULL, y = NULL) + theme_q3()

p1 + p2
```

60-month loans default more than 36-month ones. Renters default a bit more than homeowners, but the gap is small.

## Income and DTI

```{r income-dti, fig.height=4}
p3 <- rate_tables$income_band %>%
    filter(n_loans >= 50) %>%
    ggplot(aes(x = income_band, y = default_rate)) +
    geom_col(fill = "#27AE60", alpha = 0.8) +
    geom_text(aes(label = sprintf("%.1f%%", default_rate*100)), vjust = -0.5, size = 3) +
    scale_y_continuous(labels = percent, expand = expansion(mult = c(0,0.12))) +
    labs(title = "By income band", x = NULL, y = "Default Rate") + theme_q3()

p4 <- rate_tables$dti_band %>%
    filter(n_loans >= 50) %>%
    ggplot(aes(x = dti_band, y = default_rate)) +
    geom_col(fill = "#E67E22", alpha = 0.8) +
    geom_text(aes(label = sprintf("%.1f%%", default_rate*100)), vjust = -0.5, size = 3) +
    scale_y_continuous(labels = percent, expand = expansion(mult = c(0,0.12))) +
    labs(title = "By DTI band", x = NULL, y = NULL) + theme_q3()

p3 + p4
```

Lower income and higher DTI both push default rates up — as expected.

---

# 3. Who Defaults Most?

Putting it all together, the riskiest borrower profile looks like this:

```{r risky-profile}
tibble(
    Factor = c("Credit Grade", "Loan Purpose", "Loan Term",
               "Income", "DTI", "Home Ownership"),
    `Highest Default Risk` = c(
        "Grade F / G",
        "Small business loans",
        "60-month (long term)",
        "Below $30k",
        "Above 25",
        "Renters (slightly higher)")
) %>%
    kable(format = "pipe")
```

And if we look at interest rate — borrowers paying higher rates are more likely to default:

```{r int-rate-plot, fig.height=4}
closed %>%
    mutate(rate_band = cut(int_rate, breaks = seq(4, 28, by = 3), include.lowest = TRUE)) %>%
    filter(!is.na(rate_band)) %>%
    group_by(rate_band) %>%
    summarise(default_rate = mean(is_default), n = n(), .groups = "drop") %>%
    filter(n >= 30) %>%
    ggplot(aes(x = rate_band, y = default_rate, group = 1)) +
    geom_line(linewidth = 1, color = "#2980B9") +
    geom_point(size = 2.5, color = "#2980B9") +
    scale_y_continuous(labels = percent) +
    labs(title = "Default rate by interest rate",
         subtitle = "Higher rates correlate with more defaults",
         x = "Interest Rate Range (%)", y = "Default Rate") +
    theme_q3() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# 4. DTI Hard-Cap: What Should It Be?

The Director wants a recommended Debt-to-Income (DTI) cutoff. Let's look at how default rate changes as DTI rises.

```{r dti-curve, fig.height=4.5}
dti_curve <- closed %>%
    filter(!is.na(dti)) %>%
    mutate(dti_bin = cut(dti, breaks = seq(0, 40, by = 2), include.lowest = TRUE)) %>%
    group_by(dti_bin) %>%
    summarise(default_rate = mean(is_default), n = n(), .groups = "drop") %>%
    filter(n >= 30)

dti_curve %>%
    ggplot(aes(x = dti_bin, y = default_rate, group = 1)) +
    geom_line(linewidth = 1, color = "#2980B9") +
    geom_point(size = 2.5, color = "#2980B9") +
    scale_y_continuous(labels = percent) +
    labs(title = "Default rate by DTI",
         subtitle = "Risk picks up noticeably above DTI of ~25",
         x = "DTI Range", y = "Default Rate") +
    theme_q3() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Suggested caps depending on how much risk the Institute is willing to accept:

```{r dti-caps}
tolerances <- c(0.12, 0.15, 0.18, 0.20)

dti_caps <- map_dfr(tolerances, function(tol) {
    cap <- dti_curve %>%
        filter(default_rate <= tol) %>%
        pull(dti_bin) %>%
        as.character() %>%
        str_extract("\\d+(?=\\])") %>%
        as.numeric() %>%
        max(na.rm = TRUE)
    tibble(`Risk tolerance` = sprintf("%.0f%%", tol*100),
           `Suggested DTI cap` = cap)
})

dti_caps %>% kable(format = "pipe")
```

## Does it differ by state?

```{r dti-state, fig.height=4.5}
closed %>%
    filter(!is.na(dti)) %>%
    mutate(region = if_else(addr_state == "TX", "Texas", "Other States"),
           dti_bin = cut(dti, breaks = seq(0, 40, by = 3), include.lowest = TRUE)) %>%
    group_by(dti_bin, region) %>%
    summarise(default_rate = mean(is_default), n = n(), .groups = "drop") %>%
    filter(n >= 20) %>%
    ggplot(aes(x = dti_bin, y = default_rate, color = region, group = region)) +
    geom_line(linewidth = 1) + geom_point(size = 2) +
    scale_color_manual(values = c("Texas" = "#C0392B", "Other States" = "#2980B9")) +
    scale_y_continuous(labels = percent) +
    labs(title = "DTI vs Default: Texas compared to the rest",
         x = "DTI Range", y = "Default Rate", color = NULL) +
    theme_q3() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Texas follows the same pattern as the rest of the country, so a single national DTI cap is reasonable.

---

# 5. Is Texas Different?

```{r texas-table}
closed %>%
    mutate(region = if_else(addr_state == "TX", "Texas", "Other States")) %>%
    group_by(region) %>%
    summarise(
        Loans = n(),
        `Default Rate` = sprintf("%.1f%%", mean(is_default)*100),
        `Avg Loan` = sprintf("$%s", comma(round(mean(loan_amnt)))),
        `Avg Income` = sprintf("$%s", comma(round(mean(annual_inc)))),
        `Avg DTI` = round(mean(dti, na.rm = TRUE), 1),
        `Avg Interest Rate` = sprintf("%.1f%%", mean(int_rate, na.rm = TRUE)),
        .groups = "drop"
    ) %>%
    kable(format = "pipe")
```

```{r texas-test}
tx <- closed %>% mutate(is_tx = addr_state == "TX")
ptest <- prop.test(
    x = c(sum(tx$is_default[tx$is_tx]), sum(tx$is_default[!tx$is_tx])),
    n = c(sum(tx$is_tx), sum(!tx$is_tx))
)
```

Proportion test p-value: **`r format.pval(ptest$p.value, digits = 3)`**. Texas is **`r if_else(ptest$p.value < 0.05, "statistically different", "not significantly different")`** from the national average.

## How do all states compare?

```{r state-plot, fig.height=7}
closed %>%
    group_by(addr_state) %>%
    summarise(n = n(), default_rate = mean(is_default), .groups = "drop") %>%
    filter(n >= 200) %>%
    mutate(is_tx = addr_state == "TX") %>%
    ggplot(aes(x = reorder(addr_state, default_rate), y = default_rate)) +
    geom_segment(aes(xend = reorder(addr_state, default_rate),
                     y = 0, yend = default_rate), color = "grey70") +
    geom_point(aes(color = is_tx, size = n)) +
    scale_color_manual(values = c("grey40", "#C0392B"),
                       labels = c("Other", "Texas"), name = NULL) +
    scale_size_continuous(name = "Loans", labels = comma) +
    scale_y_continuous(labels = percent) +
    coord_flip() +
    labs(title = "Default rates by state",
         subtitle = "Texas in red — states with 200+ loans shown",
         x = NULL, y = "Default Rate") +
    theme_q3()
```

---

# 6. Testing the Institute's Beliefs

The Institute holds three long-standing beliefs. Let's check them against the data.

## Belief 1: Homeowners employed 10+ years default less on short-term loans

```{r h1}
closed %>%
    filter(term == 36) %>%
    mutate(group = case_when(
        home_ownership %in% c("OWN","MORTGAGE") & emp_length >= 10 ~ "Homeowner & 10+ yrs",
        TRUE ~ "Everyone else"
    )) %>%
    group_by(group) %>%
    summarise(Loans = n(),
              `Default Rate` = sprintf("%.1f%%", mean(is_default)*100),
              .groups = "drop") %>%
    kable(format = "pipe")
```

There is a gap, but it's modest. Homeownership and long employment help a bit — they're not a silver bullet though. **Partially supported.**

## Belief 2: States differ significantly on short-term defaults

```{r h2}
h2_data <- closed %>% filter(term == 36)
h2_test <- chisq.test(table(h2_data$addr_state, h2_data$is_default),
                       simulate.p.value = TRUE)
```

Chi-squared p-value: **`r format.pval(h2_test$p.value, digits = 4)`** — this is **`r if_else(h2_test$p.value < 0.05, "significant, so yes, states do differ. Supported.", "not significant. Not supported.")`**

## Belief 3: Credit grades predict defaults well for younger borrowers

We look at borrowers with lower FICO scores (below 680) as a proxy for less established credit histories.

```{r h3, fig.height=4}
has_fico <- "fico_mid" %in% names(closed) && !all(is.na(closed$fico_mid))

if (has_fico) {
    h3 <- closed %>% filter(fico_mid < 680)
    h3_sub <- "Borrowers with FICO < 680"
} else {
    h3 <- closed
    h3_sub <- "All borrowers (FICO not available)"
}

h3 %>%
    group_by(grade) %>%
    summarise(default_rate = mean(is_default), n = n(), .groups = "drop") %>%
    ggplot(aes(x = grade, y = default_rate, fill = grade)) +
    geom_col(show.legend = FALSE) +
    geom_text(aes(label = sprintf("%.1f%%", default_rate*100)), vjust = -0.5, size = 3.5) +
    scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.12))) +
    scale_fill_brewer(palette = "YlOrRd") +
    labs(title = "Default rates by grade", subtitle = h3_sub,
         x = "Grade", y = "Default Rate") +
    theme_q3()
```

Clear step-up from A to G — credit grades do a good job separating risk. **Supported.**

## Summary of beliefs

```{r h-summary}
tibble(
    Belief = c(
        "Homeowners + 10yr employment default less (short-term)",
        "States differ on short-term defaults",
        "Credit grades predict default for younger borrowers"
    ),
    Verdict = c(
        "Partially supported — small difference",
        if_else(h2_test$p.value < 0.05, "Supported", "Not supported"),
        "Supported — clear pattern across grades"
    )
) %>% kable(format = "pipe")
```

---

# 7. Recommendations for the Director

Based on the analysis, here are the main takeaways:

**On DTI caps:**
A DTI hard-cap between 25 and 30 makes sense. Above 30, defaults rise sharply. A single national cap works — no need for state-specific thresholds.

**On who defaults:**
The riskiest borrowers tend to have lower credit grades (D–G), take out small business loans, borrow for 60 months, and have high DTI or low income. These factors matter more than home ownership or employment length.

**On credit grades:**
The grading system works well. But it's even better when paired with DTI and FICO checks, especially for borderline grades like C and D.

**On interest rates:**
Borrowers paying very high interest rates are the most likely to default. It's worth asking whether charging risky borrowers more might actually make them more likely to fail.

**On Texas:**
Texas is broadly in line with national patterns — no special treatment needed.

**On the Institute's beliefs:**
Two out of three hold up. Homeownership and employment length help, but less than the Institute thinks. Credit grades and state-level differences are real.
